---
title: "Final Report SP19 Time Series"
author: "Colin Thornburg"
date: "April 28, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
#Final Paper

##Data Set for package: astsa - U.S. Monthly Live Births
###Data Description: 


This dataset is provided under the ASTSA package.  It contains a total of 373 observations and is a Time Series object that contains the monthly live births (adjusted) in thousands in the US from 1948 - 1979.  

see <https://www.rdocumentation.org/packages/astsa/versions/1.8/topics/birth>.


Some summary statistics, class, frequency, start/end of the dataset, and the variation (The variance is a numerical measure of how the data values is dispersed around the mean)
```{r birth}
library(astsa)
summary(birth)
class(birth)
mean(birth)
sd(birth)
min(birth)
max(birth)
frequency(birth)
start(birth)
end(birth)
```

## birth Basic Plot

Basic Plot of birth time series data:

```{r birthplot, echo=FALSE}
plot(birth)
```

##Data Exploration

Histogram

```{r birthhist, echo=FALSE}
hist(birth)
```

Boxplot across months
```{r birthBox, echo=FALSE}
boxplot(birth~cycle(birth))
```

TS Plot with abline
```{r birthabline, echo=FALSE}
time <- time(birth)
plot(birth); abline(reg=lm(birth~time)) # fit a trend line
```
The overall trend of live birth is decreasing from 1949 - 1979.  Live births seem to have some seasonality during the late summer/early fall months.  In order to remove this seasonality component, the data will need to be de-trended by using a moving average.  A log transformation does not appear to be needed as variance does not appear to increase with time.

```{r birthdetrend, echo=FALSE}
library(forecast)
trend_birth <- ma(birth, order=12)
detrend_birth <- birth - trend_birth
plot(as.ts(detrend_birth))
```

##Data Decomposition

```{r birthdecomp, echo=FALSE}
decompose_birth_add <- decompose(birth)
plot(decompose_birth_add)
```

Overall trend appears to be decreasing over time as stated above previously.  Seasonality does seem to exist in the late summer/early fall months.  Residuals appear to be random and are not significantly increasing over time.

##Regression

```{r birthregression, echo=FALSE}
birth.lm <- lm(birth ~ time)
plot(birth.lm)
coef(birth.lm)
summary(birth.lm)
```

Observing the plots above, the linear model appears to fit the data well although R-squared value is low, there still might be some predictive power.

##ARMA/ARIMA
```{r birthArima, echo=FALSE}
acf(resid(birth.lm))
pacf(resid(birth.lm))
```

Observing the ACF plot, it does not appear to be stationary and seasonality exists that needs to be accounted for in the model


##Model Diagnostics
```{r birthDiagnostics, echo=FALSE}
fitARIMA <- arima(birth, order=c(0,1,2), seasonal = list(order = c(1,1,1), period = 12),method="ML")
fitARIMA
```

```{r birthforecast, echo=FALSE}
plot(forecast(fitARIMA,h=20))
```

```{r birthResiduals, echo=FALSE}
plot(resid(fitARIMA))
acf(resid(fitARIMA))
```

Based on the plot of residuals, the mean is around 0 and do not appears to be correlated.  Looking at the ACF plot for residuals, the values are low and within the blue boundary lines.

